---
title: "STAT_306_Project_Group_10"
format: pdf
editor: visual
execute:
  echo: false # Hide code in document
---

# STAT 306 Project

## Nicholas Tam, Ivy Cheung, Kaichi Nakajima, Kevin Liu (Group 10)

## Introduction

Breast invasive ductal carcinoma (IDC) is the most common type of breast cancer, with about 80% of all forms of breast cancer being IDC, according to the American Cancer Society (DePolo, 2024). There are numerous nonsurgical treatments of IDC, such as radiotherapy, chemotherapy, and hormone therapy (Wright, 2023), and each their effectiveness is partly determined by the patient's condition, such as age and tumor stage. For instance, due to interactions between other treatments or conditions as a consequence of aging (e.g. Diabetes, liver disease, metabolism), more optimal doses of chemotherapy generally being discouraged for older patients due to potentially toxic side effects, implying a smaller difference in survival between older patients with or without chemotherapy (Given, Given, 2008). However, it is unclear how combinations of treatments can interact in a model to predict a patient’s survival until death.

For our research project, we have selected a dataset of approximately 1900 primary breast cancer samples, obtained from the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) database through cBioPortal for Cancer Genomics.

Our project question is:

**"How do radiotherapy, chemotherapy, and hormone therapy influence the length of time a patient with IDC will survive, given control variables age, surgery type, tumor stage, and their present survival status?"**

Our analysis will involve the inference of covariates within linear models, as we seek to determine the interaction of cancer therapies on allowing patients to survive longer.

-   Variables from columns 32 to 693 consist of genetic attributes containing m-RNA levels z-score for 331 genes, and mutation for 175 genes; they have been omitted due to being difficult to interpret.
-   Due to the distribution of "cancer_type_detailed" categories and for ease of computation, we will be filtering the dataset for IDC patients that are either alive or dead from disease, as IDC consists of the majority of the dataset, and the patients that have died from non-disease causes are not relevant to the project question and too nonspecific to be consistent.
-   "tumor_size" has been removed due to already partly contributing to "tumor_stage". (Susan G. Komen®, 2024)

## Analysis

### Loading relevant libraries

```{r}
library(dplyr) # Data manipulation operations
library(tidyverse) # Better presentation of data
library(ggplot2) # Provides commands to create complex plots
# install.packages("fastDummies")
library(fastDummies) 
# install.packages('stringi')
library(car) # Applied regression tools, including VIF
# install.packages('gridExtra')
library(gridExtra) # Extensions for grid system
```

### Uploading relevant table and cleaning data

-   All rows with "N/A" or "" for any relevant columns have been removed to ensure no data is missing.

-   All categorical variables are transformed to be treated as categorical by the table.

-   After initial analysis, "tumor_stage==0" and "tumor_stage==4" have been removed due to lack of sufficient amounts of data points.

```{r}
cancer = read.csv("METABRIC_RNA_Mutation.csv", header=TRUE)[,1:31]
cancer_new_1 <- cancer %>% 
  drop_na(age_at_diagnosis, type_of_breast_surgery, tumor_size, tumor_stage, chemotherapy, hormone_therapy, radio_therapy, overall_survival_months, death_from_cancer) %>% 
  # filter(!(age_at_diagnosis==""), !(type_of_breast_surgery==""), !(tumor_size==""), !(tumor_stage==""), !(chemotherapy==""), !(hormone_therapy==""), !(radio_therapy==""), !(overall_survival_months==""), !(death_from_cancer==""), cancer_type_detailed == "Breast Invasive Ductal Carcinoma",) %>% 
  filter(!(age_at_diagnosis==""), !(type_of_breast_surgery==""), !(tumor_size==""), !(tumor_stage==""), !(chemotherapy==""), !(hormone_therapy==""), !(radio_therapy==""), !(overall_survival_months==""), !(death_from_cancer==""), !(death_from_cancer=="Died of Other Causes"), cancer_type_detailed == "Breast Invasive Ductal Carcinoma",) %>% 
  transform(
    type_of_breast_surgery = as.factor(type_of_breast_surgery), 
    tumor_stage = as.factor(tumor_stage), 
    chemotherapy = as.factor(chemotherapy), 
    hormone_therapy = as.factor(hormone_therapy), 
    radio_therapy = as.factor(radio_therapy), 
    death_from_cancer = as.factor(death_from_cancer)
    ) %>% 
  select(age_at_diagnosis, type_of_breast_surgery, tumor_stage, chemotherapy, hormone_therapy, radio_therapy, death_from_cancer, overall_survival_months) 
# summary(cancer_new_1)

cancer_new <- cancer_new_1 %>% 
  filter(!(tumor_stage==0), !(tumor_stage==4),)
  
cancer_new$tumor_stage <- relevel(factor(cancer_new$tumor_stage), ref = "1")
cancer_new$death_from_cancer <- relevel(factor(cancer_new$death_from_cancer), ref = "Living")

head(cancer_new, 10)

summary(cancer_new)
```

### Exploratory data analysis

#### Residual plot and QQ-plot

-   There does not appear to be any observable patterns in residual values beyond being centered around 0.

-   QQ-plot indicates distribution of residuals is light-tailed normal.

```{r}
reg_start = lm(overall_survival_months~age_at_diagnosis,data=cancer_new)
plot(reg_start$fitted.values, reg_start$residuals, xlab="Fitted Values", ylab="Residuals")
abline(a = 0, b = 0)
lines(lowess(reg_start$fitted.values, reg_start$residuals), col = "blue", lwd = 2)
qqnorm(reg_start$residuals)
qqline(reg_start$residuals)
```

#### **Boxplots of survival against interactions between age and categorical variables**

-   overall_survival_months for those that took chemotherapy is generally lower than for those without in terms of mean and median, and the range of overall_survival_months is smaller.

-   Mean and median overall_survival_months for those that took hormone therapy are lower than those without, while the interquartile range for patients with hormone therapy is smaller than for those without.

-   Boxplots for overall_survival_months against each radio_therapy class are extremely similar, including the mean and median values.

-   overall_survival_months for those with mastectomy is lower than for those with breast conserving surgery in terms of mean and median, and the values are more skewed toward lower values.

-   Increasing levels of tumor stages provide decreasing mean, median and percentile values.

-   Patients that died from disease have significantly lower overall_survival_months values than those who lived and the most extreme outliers.

```{r}

box_age_chemo <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(chemotherapy, overall_survival_months, fill = chemotherapy)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
#        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for Chemotherapy",
       x = "Chemotherapy",
       y = "Overall Survival (Months)", 
       color = "Chemotherapy") + 
    stat_summary(aes(chemotherapy, overall_survival_months, fill = chemotherapy),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

box_age_hormo <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(hormone_therapy, overall_survival_months, fill = hormone_therapy)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
#        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for Hormone Therapy",
       x = "Hormone Therapy",
       y = "Overall Survival (Months)", 
       color = "Hormone Therapy") + 
    stat_summary(aes(hormone_therapy, overall_survival_months, fill = hormone_therapy),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

box_age_radio <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(radio_therapy, overall_survival_months, fill = radio_therapy)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
#        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for Radio Therapy",
       x = "Radio Therapy",
       y = "Overall Survival (Months)", 
       color = "Radio Therapy") + 
    stat_summary(aes(radio_therapy, overall_survival_months, fill = radio_therapy),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

box_age_type <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(type_of_breast_surgery, overall_survival_months, fill = type_of_breast_surgery)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
        axis.text.x = element_text(angle = 45)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for each Breast Surgery",
       x = "Breast Surgery",
       y = "Overall Survival (Months)", 
       color = "Breast Surgery") + 
    stat_summary(aes(type_of_breast_surgery, overall_survival_months, fill = type_of_breast_surgery),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

box_age_tumor <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(tumor_stage, overall_survival_months, fill = tumor_stage)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
#        axis.text.x = element_text(angle = 45)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for Tumor Stage",
       x = "Tumor Stage",
       y = "Overall Survival (Months)", 
       color = "Tumor Stage") + 
    stat_summary(aes(tumor_stage, overall_survival_months, fill = tumor_stage),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

box_age_death <- cancer_new %>%
    ggplot() +
    geom_boxplot(aes(death_from_cancer, overall_survival_months, fill = death_from_cancer)) +
    theme(
        text = element_text(size = 6),
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"), 
        axis.text.x = element_text(angle = 45)
    ) +
    ggtitle("Age Comparison") +
    labs(title = "Overall Survival for Living Condition",
       x = "Living Condition",
       y = "Overall Survival (Months)", 
       color = "Living Condition") + 
    stat_summary(aes(death_from_cancer, overall_survival_months, fill = death_from_cancer),
                 fun = mean, colour = "yellow", geom = "point", shape = 18, size = 5
                )

# grid.arrange(box_age_chemo, box_age_hormo, nrow = 1)
# grid.arrange(box_age_radio, box_age_type, nrow = 1)
# grid.arrange(box_age_tumor, box_age_death, nrow = 1)

grid.arrange(box_age_chemo, box_age_hormo, box_age_radio, nrow = 1)
grid.arrange(box_age_type, box_age_tumor, box_age_death, nrow = 1)
```

#### Scatterplots of survival against interactions between age and categorical variables

-   overall_survival_months varies significantly by chemotherapy, tumor_stage, and death_from_cancer, and the interaction terms for age with chemotherapy, type_of_breast_surgery, tumor_stage, and death_from_cancer appear to be statistically significant.

-   Confidence intervals of linear models heavily overlap for hormone_therapy and radio_therapy. 

-   Confidence intervals of linear models for chemotherapy, type_of_breast_surgery and tumor_stage only overlap for more extreme age values.

```{r}
options(repr.plot.width = 20, repr.plot.height = 5)

reg_chemo = lm(overall_survival_months~age_at_diagnosis*chemotherapy,data=cancer_new)  # TODO
summary(reg_chemo)
plot_age_chemo <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = chemotherapy)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Chemotherapy",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Chemotherapy")

reg_hormo = lm(overall_survival_months~age_at_diagnosis*hormone_therapy,data=cancer_new)  # TODO
summary(reg_hormo)
plot_age_hormo <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = hormone_therapy)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Hormone Therapy",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Hormone Therapy")

reg_type = lm(overall_survival_months~age_at_diagnosis*radio_therapy,data=cancer_new)  # TODO
summary(reg_type)
plot_age_radio <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = radio_therapy)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Radio Therapy",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Radio Therapy")

reg_type = lm(overall_survival_months~age_at_diagnosis*type_of_breast_surgery,data=cancer_new)  # TODO
summary(reg_type)
plot_age_type <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = type_of_breast_surgery)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Breast Surgery",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Breast Surgery")

reg_tumor = lm(overall_survival_months~age_at_diagnosis*tumor_stage,data=cancer_new)  # TODO
summary(reg_tumor)
plot_age_tumor <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = tumor_stage)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Tumor Stage",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Tumor Stage")

# reg_3_1 = lm(overall_survival_months~age_at_diagnosis+radio_therapy,data=cancer_new)  # TODO
# summary(reg_3_1)
# scatterplot(overall_survival_months~age_at_diagnosis+radio_therapy, data=cancer_new, smooth=FALSE, main="Overall Survival against Age and Radio Therapy", xlab="Age (Years)", ylab="Overall Survival (Months)")

reg_death = lm(overall_survival_months~age_at_diagnosis*death_from_cancer,data=cancer_new)  # TODO
summary(reg_death)
plot_age_death <- ggplot(cancer_new, aes(x = age_at_diagnosis, y = overall_survival_months, color = death_from_cancer)) +
  geom_point() +  
  geom_smooth(method = "lm", se= TRUE, na.rm = TRUE) +  
  theme(
    text = element_text(size = 7), 
    plot.title = element_text(face = "bold"), 
    axis.title = element_text(face = "bold") 
  ) +
  labs(title = "Overall Survival against Age and Living Condition",
       x = "Age (Years)",
       y = "Overall Survival (Months)", 
       color = "Living Condition")

# grid.arrange(plot_age_chemo, plot_age_hormo, plot_age_radio, ncol = 1)
# grid.arrange(plot_age_type, plot_age_tumor, plot_age_death, ncol = 1)

plot_age_chemo
plot_age_hormo
plot_age_radio
plot_age_type
plot_age_tumor
plot_age_death
```

#### Histograms of survival against interactions between age and categorical variables

-   The histogram for those that took chemotherapy compared to that for those without chemotherapy is significantly more skewed to the right.

-   Histogram peaks for those that took hormone therapy are generally greater than for those without hormone therapy within the range of overall_survival_months between 45 and 195.

-   The histogram distributions between radio therapy categories are relatively similar.

-   The histogram for those that took mastectomy compared to that for those with breast preserving surgery is significantly more skewed to the right.

-   The histogram for those with stage 3 tumors compared to that for the other tumor stages is significantly more skewed to the right.

-   The histogram for those that died from disease is significantly skewed to the right, while the histogram for those that lived is significantly skewed to the left in comparison.

```{r}
histo_age_type <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = type_of_breast_surgery)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Breast Surgery",
         x = "Overall Survival (Months)",
       y = "Density",
         color = "Breast Surgery")

histo_age_chemo <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = chemotherapy)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Chemotherapy",
       x = "Overall Survival (Months)",
       y = "Density",
       color = "Chemotherapy")

histo_age_hormo <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = hormone_therapy)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Hormone Therapy",
       x = "Overall Survival (Months)",
       y = "Density",
       color = "Hormone Therapy")

histo_age_radio <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = radio_therapy)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Radio Therapy",
       x = "Overall Survival (Months)",
       y = "Density",
       color = "Radio Therapy")

histo_age_tumor <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = tumor_stage)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Tumor Stage",
       x = "Overall Survival (Months)",
       y = "Density",
       color = "Tumor Stage")

histo_age_death <- ggplot(cancer_new, aes(x = overall_survival_months, y = after_stat(density), fill = death_from_cancer)) + 
    geom_histogram(alpha=0.5, position="identity", binwidth = 5,) +
    # facet_wrap(~location, scales = "fixed", nrow = 5) +
    theme(
      text = element_text(size = 7), 
      plot.title = element_text(face = "bold"), 
      axis.title = element_text(face = "bold") 
    ) +
    labs(title = "Overall Survival against Age and Living Condition",
       x = "Overall Survival (Months)",
       y = "Density", 
       color = "Living Condition")

# grid.arrange(histo_age_chemo, histo_age_hormo, histo_age_radio, ncol = 1)
# grid.arrange(histo_age_type, histo_age_tumor, histo_age_death, ncol = 1)

histo_age_chemo
histo_age_hormo
histo_age_radio
histo_age_type
histo_age_tumor
histo_age_death
```

### Model selection

-   As our goal is to determine how the various forms of therapy influence a patient's survival until death, interaction terms regarding all combinations of therapies will be considered, since it is unclear if the effect of one therapy will influence the effect of another (e.g. Chemotherapy and hormone therapy in similar timeframe).

-   Interaction terms for age and each type of therapy are included, as prior studies have indicated varying degrees of influence between age and treatment method (Given, Given 2008; Cleveland Clinic, 2024; U.S. National Library of Medicine; Steinfeld, Diamond, Hanks, Coia, Kramer, 1989).

-   Interaction terms for tumor_stage and each type of therapy are also included since prior studies have indicated that the type of treatment a patient receives is influenced by the stage, size of tumor and the spread of cancer cells. (“Invasive Ductal Carcinoma”, 2024;”Treatment of breast cancer”, 2024)

-   Due to not being significant parts of the question of interest and the indeterminate interaction between them and the therapies, interaction terms for type_of_breast_surgery and death_from_cancer are ignored for the full model.

```{r}
# fullmod = lm(overall_survival_months ~ .*.,data=cancer_new)
fullformula = overall_survival_months ~ age_at_diagnosis + chemotherapy * hormone_therapy * radio_therapy + age_at_diagnosis : chemotherapy + age_at_diagnosis : hormone_therapy + age_at_diagnosis : radio_therapy + tumor_stage + tumor_stage : chemotherapy + tumor_stage : hormone_therapy + tumor_stage : radio_therapy + type_of_breast_surgery + death_from_cancer 
# + death_from_cancer : chemotherapy + death_from_cancer : hormone_therapy + death_from_cancer : radio_therapy

fullmod = lm(fullformula, data=cancer_new)
# summary(fullmod)

finalmod_both = step(fullmod, direction = "both", trace = 1)
# finalmod_both = step(fullmod, direction = "both", trace = 0)
summary(finalmod_both)
```

```{r}
finalmod_backward = step(fullmod, direction = "backward", trace = 0)
summary(finalmod_backward)
```

```{r}
nullmod = lm(overall_survival_months ~ 1, data = cancer_new)
finalmod_forward = step(nullmod, scope = list(lower = nullmod, upper = fullmod), direction = "forward", trace = 0)
summary(finalmod_forward)
```

```{r}
summary(finalmod_both)

finalmod_both_1 = lm(formula = overall_survival_months ~ age_at_diagnosis + chemotherapy + hormone_therapy + tumor_stage + death_from_cancer + chemotherapy:tumor_stage + hormone_therapy:tumor_stage, data = cancer_new)
summary(finalmod_both_1)

# install.packages("AICcmodavg")
library(AICcmodavg)
aictab(cand.set = list(finalmod_both, finalmod_both_1))
```

```{r}
library(leaps)
finalmod_reg = regsubsets(fullformula, data=cancer_new, method="exhaustive", nvmax = length(attr(terms(fullformula), "term.labels")))
summary(finalmod_reg)
```

```{r}
# Create the residual plot
# TODO
plot(finalmod_both$fitted.values, finalmod_both$residuals, xlab="Fitted Values", ylab="Residuals")
abline(a = 0, b = 0)
lines(lowess(finalmod_both$fitted.values, finalmod_both$residuals), col = "blue", lwd = 2)

plot(finalmod_both_1$fitted.values, finalmod_both$residuals, xlab="Fitted Values", ylab="Residuals")
abline(a = 0, b = 0)
lines(lowess(finalmod_both$fitted.values, finalmod_both$residuals), col = "blue", lwd = 2)
```

```{r}
qqnorm(finalmod_both$residuals)
qqline(finalmod_both$residuals)

qqnorm(finalmod_both_1$residuals)
qqline(finalmod_both_1$residuals)
```

## Discussion

### **Analysis**

The final regression model includes the following key variables: age, chemotherapy, hormone therapy, and tumor stage (specifically stages 2 and 3). It also accounts for the outcome of death from cancer (death from cancer and death from other causes) and incorporates interaction terms between age and chemotherapy, chemotherapy and tumor stage (for stages 2 and 3), as well as hormone therapy and tumor stage (for stages 2 and 3).

The results from the three model selection methods (bidirectional selection, backward selection, and forward selection) consistently provided the same set of covariates, and all of them are statistically significant on the 10% significance level, and only age_at_diagnosis:chemotherapy1, chemotherapy1:tumor_stage3, and hormone_therapy1:tumor_stage2 are not statistically significant on the 5% significance level. This consistency across methods enhances the credibility of the findings, indicating that the observed relationships are relatively reliable and not due to random chance.

In the model selection, we discarded radiotherapy, indicating that it doesn’t provide a significant influence on the survival period of patients. For chemotherapy, the interaction terms between chemotherapy: tumor stage 2 and chemotherapy: tumor stage 3 provide insight into the relationship between tumor stage and chemotherapy. In tumor stage 1, the slope of -62.7677 shows the most negative correlation with survival. This negative effect diminishes in stage 2 with a slope of -69.5494, and further decreases in stage 3 with a slope of -60.5349. This indicates that while chemotherapy consistently reduces survival periods, its negative effects on survival are lower at higher tumor stages. 

For hormone therapy, it initially reduces the survival period by 24.6917 units at tumor stage 1, but its effect decreases to a reduction of only 5.9808 units in stage 2, and it even leads to an increase in survival by 19.9754 units in stage 3. From tumor stage 1 to stage 3, the impact of hormone therapy changes from harmful to beneficial, ultimately improving survival chances. This highlights the importance of carefully considering the use of hormone therapy based on tumor stage.

To further validate the reliability of the model, the plot of fitted values versus residuals shows a good fit with residuals centered around zero, demonstrating homoscedasticity. Additionally, the residuals are randomly scattered without any discernible pattern beyond being centered around 0, suggesting that the errors are independent and that the relationship between the independent variables and the dependent variable is assumed to be linear. Furthermore, the points in the QQ plot align closely along the diagonal line, indicating that the residuals are approximately normally distributed. These observations confirm that the regression model does not violate the key assumptions of regression analysis, thereby minimizing the risk of drawing incorrect conclusions.

### **Conclusion** 

To address our initial research question, our model suggests that chemotherapy and hormone therapy have varying influences on survival time before death depending on the initial tumor stage while radiotherapy does not have a significant association with survival time before death. For chemotherapy,

## Appendix

-   Breast cancer gene expression profiles (METABRIC). Kaggle. (2016, May 10). <https://www.kaggle.com/datasets/raghadalharbi/breast-cancer-gene-expression-profiles-metabric>

-   CBioPortal for Cancer Genomics. (n.d.). [https://www.cbioportal.org/study/summary?id=brca_metabric](#0)

-   DePolo, J. (2024, October 2). Invasive ductal carcinoma (IDC). Breastcancer.org - Breast Cancer Information and Support. <https://www.breastcancer.org/types/invasive-ductal-carcinoma>

-   Wright , P. (2023, March 21). Invasive ductal carcinoma (IDC). Johns Hopkins Medicine. <https://www.hopkinsmedicine.org/health/conditions-and-diseases/breast-cancer/invasive-ductal-carcinoma-idc#:~:text=Radiation%20therapy%20might%20be%20part,lymph%20nodes%2C%E2%80%9D%20Wright%20says>. 

-   Tumor size and staging. Susan G. Komen®. (2024, May 2). <https://www.komen.org/breast-cancer/diagnosis/stages-staging/tumor-size/#:~:text=Tumor%20size%20is%20related%20to,the%20size%20of%20the%20tumor>. 

-   Given, B., & Given, C. W. (2008, December 15). Older adults and cancer treatment. Cancer. <https://pmc.ncbi.nlm.nih.gov/articles/PMC2606910/#S10>

-   Hormone therapy for cancer. Cleveland Clinic. (2024, May 1).<https://my.clevelandclinic.org/health/treatments/17108-hormone-therapy-to-treat-cancer> 

-   U.S. National Library of Medicine. (n.d.). Aging changes in hormone production: Medlineplus medical encyclopedia. MedlinePlus.<https://medlineplus.gov/ency/article/004000.htm#:~:text=In%20women%2C%20estrogen%20and%20prolactin,Cortisol> 

-   Steinfeld, A. D., Diamond, J. J., Hanks, G. E., Coia, L. R., & Kramer, S. (1989). Patient age as a factor in radiotherapy. Data from the patterns of care study. Journal of the American Geriatrics Society, 37(4), 335–338.<https://doi.org/10.1111/j.1532-5415.1989.tb05501.x>
